In-silico molecular enrichment and clearance of the human intracranial space

### 2024-11-27 paper/main.tex
Clean-up and shuffle.
Former-commit-id: 5fdf3caa5f44975b8abfa821b108e6866cf9bb64

 \documentclass[fleqn,10pt]{wlscirep}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{bm}
\usepackage{caption}
\usepackage{subcaption}
%\usepackage{natbib}
\graphicspath{{figures/}}

\title{Molecular transport in human pial perivasculature}

\author[1,x]{Rami Masri}
\author[1,x]{Miroslav Kuchta}
\author[1,x]{Marius Causemann}
\author[1,*]{Marie E. Rognes}
\affil[1]{Department of Numerical Analysis and Scientific Computing, Simula Research Laboratory, Oslo, Norway}
\affil[x]{Author order to be discussed.}
\affil[*]{meg@simula.no}

\newcommand{\rami}[1]{\textcolor{blue}{#1}}

@@

@@

\begin{itemize}
\item Vinje et al~\cite{vinje2021brain} and references therein illustrates the location and characteristics of human pial perivascular spaces.
\item Mestre et al~\cite{mestre2022periarteriolar} study the properties of pial perivascular spaces in detail in mice.
\item Mestre et al~\cite{mestre2018flow} measure and characterize pial perivascular transport and estimate flow magnitudes (in mice).
\item
  Weller and coauthors conducted pioneering studies of perivascular spaces, flow and transport~\cite{zhang1990interrelationships, zhang1992directional}. They identified thin sheaths of pial cells surrounding arteries and arterioles (but not veins or venules) on the brain surface and within the human brain itself~\cite{zhang1990interrelationships}. Further, they observed that tracers spread along perivascular (arterial, venous and capillar) spaces in (rat) grey matter, and in the subarachnoid space (to the cribriform plate and nasal lymphatics)..   
\item In a series of papers, Bakker and coauthors study perivascular anatomy and solute transport~\cite{bedussi2017paravascular, bedussi2018paravascular}. They find that the subarachnoid space, the cisterns, ventricles and penetrating periarteriolar spaces form a continuous cerebrospinal-fluid filled space surrounding and penetrating the rat brain\cite{bedussi2017paravascular}. Moreover, they demonstrate pulsatile and directional (antegrade) flow in perivascular (predominantly periarterial) spaces on the brain surface.
\item
  The architecture of pial and subarachnoid perivascular spaces in the rat brain was also studied by Thorne and colleagues~\cite{pizzo2018intrathecal, hannocks2018molecular}.
\item Bollman et al~\cite{bollmann2022imaging} image the human pial vasculature at high resolution (7T).
\item Hornkjøl et al~\cite{hornkjol2022csf} model CSF flow and solute transport in the human SAS and brain parenchyma. 
\item Mardal et al~\cite{mardal2022mathematical} give a succinct introduction to MRI-based computational brain modelling. 
\item
  Rey et al~\cite{rey2023perivascular} (from the group of Sartinoranont) model and study perivascular transport in rat brain networks.
\end{itemize}

\section*{Methods}

\subsection*{Meshing}

Hodneland et al~\cite{hodneland2019new} developed a method to image the brains' microvasculature based on time-of-flight (ToF) and quantitative susceptibility mapping (QSM) to identify arterial and venous networks.

We derive a detailed tetrahedral mesh of both the parenchyma and the surrounding CSF-filled spaces alongside a network representation of the vasculature from the anatomical data provided therein.

\paragraph{3D domain}

We merge the white and grey matter masks provided by \cite{hodneland2019new} to extract the pial surface. As no data on the CSF-filled spaces is available, we smooth and extend the pial surface by about 5\,mm to obtain an approximation of the SAS and include the inner fluid-filled cavities such as the ventricular system. 
Finally, we generate a tetrahedral mesh of both compartments using the fTetWild meshing algorithm \cite{hu2020fast}.

\paragraph{Vessel networks}

We skeletonize the binary masks of both the arterial and venous network using Kiminaro \cite{william_silversmith_2021_5539913} and obtain networks consisting of centerline and radius information of all detected vessels. The resulting networks are further smoothened using a spline interpolation technique and the nodes with lowest z-coordinate are identified as the networks' root nodes.

\subsection*{Geometries and domains}

The computational domain $\Omega$ is defined by the union of the SAS $\Omega_{\rm sas}$ and brain parenchyma $\Omega_{\rm brain}$. The thin pia membrane encloses the brain, and thus the pial surface (of the brain) defines the SAS-brain interface $\Gamma_{\rm pia}$. A network of blood vessels are embedded in the SAS, potentially very near to or on the pial surface. We assume that a network of idealized (annular or semi-annular) perivascular spaces surround the vasculature. We represent the vasculature and associated perivasculature by a network of one-dimensional curves $\Gamma$ representing the network centerlines equipped with inner $R_1$ and outer $R_2$ radius information.

Surface representations of the outer SAS boundary, pial membrane, and delimiters of other CSF-filled spaces such as the lateral ventricles, third and fourth ventricle, and aqueduct are extracted from MR-images of human subject(s), see e.g. the pipeline described by Hornkjøl et al.~\cite{hornkjol2022csf}. From these surface representations, we create volumetric meshes of the SAS and brain using SVMTK~\cite{mardal2022mathematical}.

\begin{itemize}
    \item To discuss: How to generate subject-specific representations of the pial/SAS vasculature? Can we use contrast MRI? Can we use time-of-flight images? What are the right tools? Do any data sets already exist? 
    \item How to add radius information to vasculature representation? Do we need more information as well?
\end{itemize}

\begin{figure}
\caption{(A) Illustration of the location of pial vasculature and subarachnoid space. (B) Conceptual illustration of the vascular, perivascular, SAS and cortex domains (cross-section and lateral view).}
\label{fig:concept}
\end{figure}

\subsection*{Multi-dimensional transport model}

We consider a time-dependent convection-diffusion model~\cite{masri2023modelling} of a solute concentration $c$ in the network representing the pial vasculature and surrounding perivascular spaces $\Lambda$ and the surrounding three-dimensional SAS-brain domain $\Omega$ (Fig.~\ref{fig:concept}).

\subsection*{Diffusion}
We use the diffusion coefficient $D_{\rm gad}$ of Gadubutrol in free water in the SAS and PVS, and effective diffusion coefficient of Gadubutrol $D_{\rm brain}$ -- accounting for the extracellular space (ECS) porosity $\phi$ and the ECS tortuousity $\lambda$ but ignoring anisotropy  -- in the brain parenchyma~\cite{hornkjol2022csf}.

We solve for two concentration fields $c, \hat c$ in the domain $\Omega$ and in the network $\Lambda$. The 
network $\Lambda$ consists of 1D centerlines of vessels with their radii information and nodes. We consider the following model. 
\begin{alignat}{2}
\partial_t c - \nabla \cdot (D \nabla c ) + \xi (\overline{c} - \hat c ) \delta_\Gamma & = f, && \quad \mathrm{in } \,\, \Omega \\ 
\partial_t (A\hat c) - \partial_s(\hat D A  \partial_s \hat c) + \xi P (\hat c - \overline{c})   & = A \hat f, && \quad \mathrm{in } \,\,  \Lambda 
 \end{alignat}
We split the boundary of $\Omega$ into $\Gamma_{\mathrm{skull}}$ and $\Gamma_{\mathrm{spine}}$,  and we enforce the following boundary conditions. 
\begin{alignat}{2}
D \nabla c \cdot \bm{n} &= 0, && \quad \mathrm{on} \,\, \Gamma_{\mathrm{skull}}, \\ 
D \nabla c \cdot \bm{n} &= g(t),  && \quad \mathrm{on} \,\, \Gamma_{\mathrm{spine}}.
\end{alignat}
The function $g$ is given in 
\begin{equation*}
    g(t) = \max(0, a(T_0 - t))
\end{equation*}
For the network, the boundary nodes are split into $\partial \Lambda^{\mathrm{in}}$  and $\partial \Lambda^{\mathrm{out}}$ based on the specified orientation of the network (\color{blue} to be specified more precisely (with a figure?)\color{black}). We impose the following conditions. 
\begin{alignat}{2}
D A \partial_s \hat c &= 0, && \quad \mathrm{on} \,\, \partial \Lambda^ {\mathrm{in}}, \\ 
D A \partial_s  \hat c &= \hat g(t),  && \quad \mathrm{on} \,\, \partial \Lambda^{\mathrm{out}} .
\end{alignat}
\color{blue} Running different models depending on the choices of $g$ and $\hat g$ \color{black}
 \subsection*{Convection}
We ignore the deformation of the domains (in this study), and thus set the domain velocity $w = 0$. We assume no significant convection in the brain $\Omega_{\rm brain}$, and set $u = u_{\rm brain} = 0$ there as a first approximation. We may consider non-zero convection in the SAS $u_{\rm sas} \not = 0$ e.g.~resulting from steady state CSF production as predicted by~\cite{hornkjol2022csf} in model variations, but may start with $u_{\rm sas} = 0$ as a baseline. In the perivascular network, we set an average axial velocity $u_{\rm pvs} = \langle u_{v, s} \rangle$ of $18.7$ $\mu$m/s as reported by Mestre et al.~\cite{mestre2018flow} in mouse perivascular spaces under normal physiological conditions. We may also consider a reduced perivascular velocity corresponding to hypertension in model variations.
\subsubsection*{Steady state PVS flow model} We solve 1D Darcy equations in the vessel network. The unknowns are the averaged PVS velocity, $u_{\rm pvs} = \langle u_{v,s} \rangle$, and the cross-section pressure $p_{\rm pvs} $ (assumed to be constant in each cross-section). The system is given by  \cite{daversin2022geometrically} 
\begin{alignat}{2}
A \,  u_{\rm pvs}   + \frac{\kappa}{\mu} \, A \, \partial_{s} p_{\rm pvs} & = 0, &&  \quad \rm in  \,\, \Lambda  \\ 
-\partial_s (A \, u_{\rm pvs}) & = f, && \quad \rm in  \,\, \Lambda .  
\end{alignat} 
In the above, $A$ is the area of the PVS, $\mu$ is the dynamic viscosity, and $\kappa$ is derived from the assumptio of Poiseuille
flow in the annular cross-section of the PVS \cite{daversin2022geometrically,tithof2022network}: 
\begin{equation}
\kappa = \frac18 \left( R_2^2 + R_1^2 - \frac{1}{\ln(R_2/R_1)} (R_2^2- R_1^2) \right). 
\end{equation}
On bifurcations, conservation of fluxes is enforced weakly and continuity of the pressure is enforced strongly (in the choice of the finite element spaces).

\subsubsection*{Case 1} We prescribe zero Dirichlet boundary conditions for the pressure, and vary $f$ \rami{include the reasoning behind having this $f$ as a source or sink term, and include the value that we opted for} to obtain a flow field with a similar magnititude to what is commonly reported in the literature.

\subsection*{Case 2} We prescribe a Robin type boundary condition on the outlet nodes. In particular, we set 
\begin{equation}
A  u_{\rm pvs} =  - \frac{\kappa}{ \mu} A\, \partial_s p_{\rm pvs} = \alpha (p_{\rm pvs} - p_0), \quad  \rm on \,\,\, \partial \Lambda^{\rm out} 
\end{equation}
On $\partial \Lambda^{\rm in}$, we prescribe homogenous Dirichlet data for the pressure (\rami{does this make sense?}). 
\rami{to be completed }
\subsection*{PVS permeability}
We will consider different values for the SAS-PVS permeability $\xi$, including $\xi = 0$ (sealed PVS), $\xi \approx \infty$ (PVS continuous with SAS), as well as reasonable intermediate values. We will ignore PVS-blood exchange, and thus only model the perivascular concentration at the network level (in addition to the SAS/brain concentration in the surroundings).

\subsection*{Boundary and initial conditions}
We set no flux on all boundaries unless otherwise specified. We start with an initial non-zero concentration in the first branch of the perivascular network and surrounding SAS (and zero concentration elsewhere).

\subsection*{Parameter values}

\begin{table}
    \centering
    \begin{tabular}{|c|c|c|c|c|} \hline 
 Parameter& Symbol& Value& Unit&Reference\\ \hline 
         parenchyma app. diffusion&  $d_p$&  $1.2 \times 10^{-4}$& $\text{mm}^2/\text{s}$  & \cite{valnes2020apparent}\\\hline 
         CSF app. diffusion&  $d_{CSF}$&  $3.8 \times 10^{-4}$& $\text{mm}^2/\text{s}$ & \cite{valnes2020apparent}\\\hline 
         PVS diffusion&  $d_{PVS}$&  $3.8 \times 10^{-4}$& $\text{mm}^2/\text{s}$ & \cite{valnes2020apparent}\\\hline 
         extracellular space volume fraction& $\phi$& 0.2& - &\cite{nicholson1981ion} \\\hline 
         PVS-parenchyma permeability&  $\xi_{PVS,P}$& $5 \times 10^{2} - 6 \times 10^{3} $  & $\text{m}^{-1}$ & \cite{koch2023estimates} \\\hline 
         PVS-CSF permeability&  $\xi_{PVS,CSF}$& $\{0, \text{inf}\}$ & $\text{m}^{-1}$ & \\ \hline 
 boundary concentration& $\hat{c}_D$& & &\\ \hline
    \end{tabular}
    \caption{Caption}
    \label{tab:parameters}
\end{table}

\subsection*{Quantities of interest}

\begin{itemize}
    \item Snapshots (e.g. coronal plane) of concentration at relevant times (15 min, 30 min, 1h, 2h, ...) 
    \item Total amount of tracer in the PVS, SAS and parenchyma
     \item Time to reach a threshold (e.g. 30\%) in different "places"
\end{itemize}

\section*{Results}

\begin{table}
\begin{center}
    \begin{tabular}{l|cc}
    \toprule
    Label & Description & Parameter values \\
    Model 0 & Baseline &  \\
    \bottomrule
    \end{tabular}
    \end{center}
    \caption{Overview of models}
\end{table}

\begin{figure}
     \centering
     \begin{subfigure}[b]{0.33\textwidth}
         \centering
         \includegraphics[width=\textwidth]{modelA_total_conc.png}
         \caption{arterial inflow, pure diffusion}
         \label{fig:y equals x}
     \end{subfigure}
     \hfill
     \begin{subfigure}[b]{0.33\textwidth}
         \centering
         \includegraphics[width=\textwidth]{modelB_total_conc.png}
         \caption{SAS inflow, pure diffusion}
         \label{fig:three sin x}
     \end{subfigure}
     \hfill
     \begin{subfigure}[b]{0.33\textwidth}
         \centering
         \includegraphics[width=\textwidth]{modelC_total_conc.png}
         \caption{arterial inflow, diffusion + PVS convection}
         \label{fig:five over x}
     \end{subfigure}
        \caption{Total tracer amount in parenchyma, CSF, arterial PVS and venous PVS}
        \label{fig:three graphs}
\end{figure}

\subsection*{Effect of interfaces (interface permeability)}

\begin{itemize}
    \item Model 0 (baseline): only diffusion, different diffusion coefficient in the parenchyma and SAS/PVS. No leak to vasculature (zero permeability). $\infty$ $\zeta_0$ at the PVS-SAS interface, $\zeta_1$ at the PVS-SAS interface. Koch et al could be a reference for PVS-parenchyma, maybe Tithof et al (2022, "A network model" ...) could give some ideas for parameters, also see Mestre et al, Nature comms, 2022. 
    \item 
    With altered interface permeabilities in the PVS (Model 1), and or in the parenchyma (Model 2) 
    \end{itemize}

\begin{figure}
    \caption{}
    \label{fig:1}
\end{figure}

\subsection*{Effect of convection}
\begin{itemize}
    \item 
    With convective flow in the PVS (zero flow in the parenchyma for now)
    \begin{itemize}
        \item 
        Model 3: PVS flow equal to 18.7 $\mu$m/s
        \item 
        Model 4: Better PVS flow (distributed, according to conservation of mass, assumptions on PVS area)
    \end{itemize}    
    \item 
    With convective flow in the ECS (out of scope here, for follow-up work?)
\end{itemize}

\begin{figure}
    \centering
    \includegraphics[width = 0.9 \textwidth]{modelB_modelC_overview.png}
    \caption{Tracer concentrations for a purely diffusive model with tracer inflow via the CSF-filled space (left) and a convective model with Tracer inflow via the arterial PVS (middle) and their difference (right) at 1, 6, 12, 18 and 24 hours after injection.}
    \label{fig:2}
\end{figure}
\subsection*{Effect of BBB}

Leakage to the blood, some sink/non-zero permeability for the PVS-blood (BBB) interface (Model N).

\begin{figure}
    \caption{}
    \label{fig:3}
\end{figure}

\subsection*{Effect of pia}

Effect of pia permeability (Model N-1)

\begin{figure}
    \caption{}
    \label{fig:4}
\end{figure}

\section*{Discussion}

@@

\begin{itemize}
    \item Pulsating brain. Natural to consider in follow-up work.
    \item PVS-blood exchange. Natural to consider in model variation here, or follow-up work.
    \item Accounting for brain diffusive anisotropy. Little effect expected. Maybe something to consider in small follow-up work.
\end{itemize}

>>>>

\documentclass[fleqn,10pt]{wlscirep}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{bm}
\usepackage{caption}
\usepackage{subcaption}
% \captionsetup[subfigure]{justification=justified,singlelinecheck=false}
\usepackage{tikz}
\usepackage{newtxmath} % MER: Prettier math
\usepackage{cleveref} % MER: Allows \Cref (but use Section X.Y not Subsection X.Y)
\usepackage{booktabs} % MER: Prettier tables

\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{float}
\usepackage{amsmath}
\usepackage{siunitx}
\usepackage{placeins}
\usetikzlibrary{quotes}
\usetikzlibrary{arrows,decorations.pathmorphing,backgrounds,positioning,fit,petri}

\graphicspath{{figures/}}

\captionsetup[subfigure]{justification=justified,singlelinecheck=false}

\title{In-silico molecular enrichment and clearance of the human intracranial space}
\author[1]{Marius Causemann}
\author[1]{Miroslav Kuchta}
\author[2]{Rami Masri}
\author[1,3,*]{Marie E. Rognes }
\affil[1]{Department of Numerical Analysis and Scientific Computing, Simula Research Laboratory, Oslo, Norway}
\affil[2]{Brown University, Providence, Rhode Island, USA}
\affil[3]{K. G. Jebsen Centre for Brain Fluid Research}
\affil[*]{meg@simula.no}

\newcommand{\rami}[1]{\textcolor{blue}{#1}}
\newcommand{\mer}[1]{\textcolor{magenta}{#1}}
\newcommand{\mar}[1]{\textcolor{violet}{#1}}
\newcommand{\discuss}[1]{\textcolor{red}{#1}}
\newcommand{\fixme}[1]{\textcolor{red}{#1}}
\newcommand{\draft}[1]{\textcolor{gray}{#1}}
%\newcommand{\brain}{\Omega_{\rm{PAR}}} 
%\newcommand{\sas}{\Omega_{\rm CSF}}
%\newcommand{\spinal}{\Gamma_{\rm SAS}}
%\newcommand{\skull}{\Gamma_{\rm{skull}}}
%\newcommand{\gin}{g_{\rm{influx}}}

\newcommand{\R}{\mathbb{R}}
\newcommand{\foralls}{\forall \,}

@@

\mer{MER: Target length: 5000 words, 4-6 figures: 600 words Intro. 1400 words Methods. 2000 words Results. 1000 words Discussion.}

@@

Motivation: as indicated by~\cite{iliff2012paravascular}, referring to
Cserr, Phys Rev., 1971, bulk convective flow may be more important for
larger species (larger distances) and larger molecules.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}
\mar{~ 2000 words and 7 subsections - around 300 words each?}
\begin{enumerate}
\item
  ``A digital twin'' for molecular enrichment and clearance in human intracranial spaces
\item
  CSF flow, the role of production versus no/cardiac/respiratory
  dispersive effects on tracer trajectories, flow bifurcation. With and without tentorium cerebellum.
\item
  ``The role'' of PVS flow speed
\item
  ``The role'' of PVS sheaths and permeability - does functional separation entail structural separation?
\item
  ``The role'' of PVS size (motivated by pathologies)
\item
  Clearance from the CSF spaces - with and without low pressure veins, but keeping the intrathecal source of solute. 
\item
  Clearance from the parenchyma: $c_0 = 1.0$. Different substances?
\end{enumerate}

\begin{figure}
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{figures/Brain-PVS-callouts.png}
     \end{subfigure}
     \hfill
     \caption{a) Illustration of the different intracranial domains
       considered with labels;}
     \label{fig:intracranial_domains}
\end{figure}

\subsection*{High-fidelity in-silico predictions of intracranial solute transport and exchange}

\begin{itemize}
\item
  Present mathematical and computational model and geometries (\Cref{tab:scenarios}). 
\item
  Highlight computational model availability, open source code.
\item
  Provides an interactive in-silico platform for studying intracranial solute transport -- distribute such that anyone can run (possibly try to recruit MinRK if advantageous).
\item
  \mer{MER: Remember to report on how much concentration enters into the brain?}
\end{itemize}

\subsection*{CSF flow and dispersion shifts enrichment patterns}

Flow of CSF through the SAS and the ventricular system is a crucial driver of transport in the human brain environment.
While cardiac- and respiratory induced oscillatory flow increases mixing, the production of CSF in the choroid plexus creates a steady advective flow within the CSF-filled spaces of the cranium \cite{hornkjol2022csf}. 
We construct a 3D model of the SAS and the ventricular system (\Cref{fig:csf}A,B and C) and compute CSF flow fields for three different scenarios: CSF production in the choroid plexus, CSF flow at systolic peak blood inflow and CSF flow in response to free breathing.

% production flow simulation
To simulate the production-driven flow, we impose an inflow of $400\,$ml/day \cite{nilsson1992circadian} across the surface of the lateral ventricle and allow a resistive outflow across the upper convexity of the arachnoid mater, leading to a total pressure drop of $26\,$mPa between the production and outflow sites (\Cref{fig:csf}D). A maximum flow velocity of $1.8\,$mm/s is attained in the aqueduct, while the mean velocity in the SAS is $2.6\,\mu$m/s (\Cref{fig:csf} E).

%The Stokes system is solved with the finite element software $FEniCS$, stored and then subsequently read for all the relevant solute transport simulations. See Figure~\ref{fig:csf} for a streamline visualization of the obtained velocity profile and \Cref{sec:details_numerical_method} for details on the numerical algorithm employed.

% cardiac pulsatility
Assessing the effect of cardiac-driven dispersion, we compute a steady CSF flow field at peak systolic blood inflow. We mimic the resulting reduction of CSF space volume by imposing an inflow of $6\,$ml/s \cite{baledent2014imaging, causemann2022human} and $0.31\,$ml/s \cite{vinje2019respiratory} across the upper arachnoid mater ($\Gamma_{\rm AM-U}$) and the lateral ventricle ($\Gamma_{\rm VL}$) surface, respectively. We allow free outflow into the spinal SAS ($\Gamma_{\rm SSAS}$). In addition, we upscale the obtained pressure field to account for the lack of pulsatile inertial effects in our model (see \ref{sec:dispersion_details}). In this scenario, we observe a pressure drop of $10\,$Pa between lateral ventricles and the spinal SAS and a maximum flow velocity of $19.8\,$cm/s in the aqueduct (\Cref{fig:csf}F and J).
Next, we estimate a spatially varying dispersion enhancement factor $R$ from cardiac-induced pulsatility by combining the computed pressure field with the theory of shear-augmented dispersion (see \Cref{sec:dispersion_details}). Assuming a cardiac frequency of $1\,$Hz, find that cardiac-induced pulsatile CSF flow increases the effective diffusion by up to two orders of magnitude in the aqueduct and around the cisterna magna, but has little effect in most of the SAS ($R < 1$) (\Cref{fig:csf} G and K).

% respiratory pulsatility
Employing the same methodology, we estimate the respiratory contribution to dispersive mixing by computing the CSF flow field driven by an inflow of $0.121\,$ml/s \cite{liu2024using} and $1\,$ml/s across the lateral ventricular wall and the arachnoid mater, respectively. This amounts to a respiratory peak flow volume of $1.121\,$ ml/s at the craniocervical junction \cite{gutierrez2022effect}. While the resulting flow velocities are only about one fourth of their cardiac-induced counterparts (see \Cref{fig:csf}H and L), respiratory dispersive mixing reaches a factor of up to 350 due the lower respiratory frequency ($0.25\,$Hz) (\Cref{fig:csf}I and M).

% tracer trajectory of the standard model
Next, we use the computed advective velcity field and dispersion enhancement factors to predict the spreading of $0.5\,$mmol intrathecally injected Gadobutrol in CSF-filled spaces, the parenchymal tissue and PVSs around arteries and veines. One hour post-injection, tracer moves upwards in the SAS frontally of the brainstem,  and quickly reaches supratentorial regions (\Cref{fig:csf}N). Here, it spreads both posteriorly through the longitudinal fissure and anteriorly through the outer SAS, reaching the top of the cerebral cortex after around 12 h. After 24h, the tracer covers most of the brains surface (with exception of some posterior regions) and has penetrated several centimeters into the parenchymal tissue.

% tracer trajectory of model variations

\begin{itemize}
    \item only dispersion model: No production flow. Tracer remains in the subtentorial regions, in cerebellum, brain stem and surrounding CSF. In contrast to other models, it travels up the ventricular system - similar to what is observed in iNPH patients?
    \item Low dispersion model: account for uncertainty (or pathologies with impaired pulsatile motion?) - tracer exclusively spreads in the frontal half of the brain. Bifurcation of CSF production flow field: flow bifurcates around the midbrain in two directions: a) through the quadrigeminal cistern into the longitudinal fissure and b) frontally around the brain through the outer SAS. Only models with "enough" dispersion enable tracer to follow trajectory a).
    \item without tentorium: 
\end{itemize}

\begin{figure}[h!]
\centering 
\includegraphics[width = 0.9\linewidth]{figures/figure2.png}
\caption{
A) Illustration of 3D-domains; 
B) Illustration of relevant boundaries; 
C) schematic illustration of the model, its doamins and boundaries; 
D) CSF flow induced by CSF production; 
E) production-induced CSF velocities in the SAS and third and fourth ventricle (V3 \& V4); 
F) CSF flow induced by peak systolic expansion; 
G) cardiac dispersion enhancement factor;
H) CSF flow induced by respiration; 
I) respiratory dispersion enhancement factor;
J) cardiac-induced flow velocities;
K) cardiac-induced dispersion enhancement factor;
L) respiratory flow velocities;
M) respiratory dispersion enhancement factor;
N) sagittal, coronal and axial planes of tracer concentration after 1, 6, 12 and 24$\,$h for model A. The outer model boundaries is shown in cyan, the pial membrane in purple and the arteries in red.;
O) model X,Y and Z after 12h;
}
\label{fig:csf}
\end{figure}

\subsection*{The significance of perivascular spaces as flow highways}
\begin{itemize}
\item
  Perivascular flow and transport has been identified as a key
  mechanism and potential target for enhancing brain drug delivery and
  metabolic waste clearance -- what are the global (intracranial)
  effects of such intracranial highways? 
\item
  Do solutes move from PVS to the SAS, or both along the PVS and along
  the SAS?
\end{itemize}
\begin{figure}
    \centering
    \includegraphics[height=\textheight]{figures/figure3.png}
    \caption{\mer{MER: Put vasculature network illustration here. Many aspects actually. Marking of segments. Location relative to 3D. Radius etc. Marius will evaluate and suggest. Results: Group segments by logical order. For key arterial segments: average radius and length (can be moved to Supplementary if need be); pressure values; flow velocities; and then histograph with flow velocities for each mesh cell. Make sure to minorly note that ACA refers to a segment of the ACA and not the whole thing e.g.}}
    \label{fig:pvs}
\end{figure}
\Cref{fig:pvs}.

\subsection*{Structural versus functional compartmentalization of perivascular spaces}

\begin{itemize}
\item
  Are perivascular spaces functionally or structurally
compartmentalized? Human and rodent observations indicate that tracers
concentrate in perivascular spaces surrounding the pial and
subarachnoid vasculature. Must these spaces be structural i.e.~defined
by semi-permeable structural
membrane~\cite{zhang1990interrelationships, zhang1992directional,
  mestre2018flow, eide2024functional}, or can such
compartmentalization be a result of regions of enhanced flow or
mixing~\cite{bedussi2017paravascular, vinje2021brain}? See also
references in Background.
\item
  Vary the PVS permeability $\xi$.
\end{itemize}

\subsection*{Enlarged perivascular spaces yield ...}

\begin{itemize}
  \item Take $R_2 = 3 R_1$ instead of $2 R_1$.
\end{itemize}

\subsection*{The perivenous contribution to molecular clearance of CSF tracers}

\begin{itemize}
\item
  Look at clearance phase 12h-24h-48h-72h.
\item
  Look at concentration in the veins
\item
  Look at concentration in the parenchyma
\item
  Consider low-pressure sink for the perivenous spaces.
\end{itemize}

\subsection*{Clearance of intraparenchymal solutes}

\begin{itemize}
\item
  Clearance of a molecular concentration in the parenchyma, different
  molecular sizes - ranging from Gadubutrol to amyloid-$\beta$/$\tau$.
\end{itemize}

\bigskip

\FloatBarrier
\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}

\paragraph{CSF flow model validation}

Hornkjøl et al \cite{hornkjol2022csf} model CSF flow induced by production in the choroid plexus and report a peak velocity of $8.9\,$mm/s in the aqueduct. Given that similar flow rates were enforced, the deviation from our value of $2.0\,$mm/s is likely due to differences in diameter of the reconstructed aqueduct. Additionally, their peak velocities in the SAS range between 20 - 50 $\mu\,$m/s, which is compatible with our mean velocity of $4.4\,\mu$m/s (\Cref{fig:csf_flow_histo}).

The aqueductal ($0.31\,$ml/s) and craniocervical ($8.31\,$ml/s) flow rates predicted by our model align well with previously reported values of $0.31\,$ml/s peak aqueductal flow rate \cite{vinje2019respiratory} and about $9\,$ml/s peak net blood inflow into the cranium \cite{baledent2014imaging}. Further, Linninger et al \cite{linninger2007cerebrospinal} model cardiac-induced CSF flow and predict a maximum transmantle pressure difference of $10\,$Pa, which is in good agreement with our estimate of $8.5\,$Pa as well as clinically measured mean pressure differences of $11.5\,$Pa between sensors placed subdurally and in the lateral ventricle \cite{vinje2019respiratory}.
% 10.7 cm/s CSF flow at the craniovertebral junction in control \cite{bapuraj2016cerebrospinal}, area difference?

Sharp et al used the theory of dispersive mixing to predict a dispersion enhancement factor of $91.8$ in the (open) spinal SAS, which lies well within the range of our spatially varying estimate (ranging from 0 to 160). Conversely, \cite{ray2021quantitative} report up to 10,000 times greater diffusion in perivascular regions, but their modelling lumps diffusive, dispersive and advective transport in a single enhancement factor and is thus not directly comparable.

@@

\begin{itemize}
\item
  \mer{MER/MC: How about dispersion in the PVS?}
\item 
  Inaccuracies in vascular geometry representation - and consequences downstream for e.g. numerical integration.
\item
  PVS geometry assumptions and approximations. Humans different from mice. Limitations in resolution.
\item
  No vasculature in the cerebellum, and of course no vasculature below certain resolution.
\item
  Only partially accounting for pulsatility. 
\item
  Accounting for brain diffusive anisotropy. Little effect expected. Maybe something to consider in small follow-up work.
\item
  Permeability and stiffness of the pial membrane
\item
  PVS-blood exchange and BBB leakage. Natural to consider in model variation here, or follow-up work. We here ignore PVS-blood exchange, and thus only model the perivascular concentration at the network level (in addition to the SAS/brain concentration in the surroundings).

\end{itemize}

\subsection*{Outlook}

\begin{itemize}
\item
  Considering the (maybe) low diffusion in CSF spaces, maybe pure
  advection would be relevant? Both computational and physiological
  implications/interest.
\end{itemize}

\newpage
\section*{Methods}

%\mer{MER: I recommend writing each (sub)section of Methods as a nearly stand-alone paragraph. Let's see if we can avoid subsubsections/paragraphs. If you need to add more detail than you think is appropriate, put it in a Supplementary Methods section in the Appendix.}

\subsection*{Intracranial compartments: ventricular system, SAS, and brain parenchyma}

From T1-weighted MR images of a human subject~\cite{hodneland2019new},
we first automatically segment the brain parenchyma and CSF spaces,
including the ventricular system and SAS, using
Synthseg~\cite{billot2023robust,billot2023synthseg}. We next manually
adjust the segmentation to accurately represent the connections
between CSF spaces (aqueduct, median aperture), smoothen, and finally
extract surface representations of the outer (arachnoid) boundary, the
pial membrane and other interfaces. Conforming to these surface and
interface representations, we generate a tetrahedral mesh $\Omega$
representing the complete intracranial volume as the union of the
parenchyma $\Omega_{\rm PAR}$ and CSF spaces $\Omega_{\rm CSF}$, using
fTetWild~\cite{hu2020fast}. The resulting computational mesh consists
of $153\,729$ vertices and $851\,904$ mesh cells, which vary between
$0.4$ mm and $1.2$ cm in mesh cell diameter. The parenchyma has a
total volume of $1386$ ml, whereas the ventricles and the outer SAS
contribute $28$ ml and $277$ ml, respectively, to a total CSF volume
of $305$ml, in agreement with recent
estimates~\cite{hladky2024regulation}.

%% In addition, we slightly modify the segmentation to ensure correct
%% connectivity of the ventricular system. In particular, we elongate
%% the aqueduct and enforce no communication between the ventricles
%% and the surrounding CSF spaces except for the median aperture.
%% After morphological smoothing of the segmented image, we extract
%% surface representations of the outer SAS boundary, the pial
%% membrane and the components of the ventricular system.

%% Note that the widely reported value of $150$\,ml of intracranial
%% CSF volume is likely an underestimate. Recent estimates indicate
%% that a value of $200$ to $300$\,ml is more accurate (see
%% \cite{hladky2024regulation} and references therein

\subsection*{Periarterial and perivenous spaces}

To represent surface networks of periarterial and perivenous spaces,
we use Kiminaro~\cite{william_silversmith_2021_5539913} to separately
skeletonize time-of-flight angiography (ToF) and quantitative
susceptibility mapping (QSM) images from the same human
subject~\cite{hodneland2019new}. This technique yields networks of
one-dimensional curves, each curve $\Lambda^i$ indicating the
centerline of a blood vessel segment and labeled with its lumen radius
$R_1^i$. We also assign an outer radius $R_2^i > R_1^i$ to each
segment $i$, form the annular cylinder ensheathing $\Lambda^i$ and
define the union of these as the PVS. The periarterial network graph
consists of $12\,708$ edges connected at $12\,562$ inner nodes and
with $147$ end nodes, and the associated domain is denoted by
$\Lambda_a$. We identify three of the end nodes -- corresponding to
the two internal carotid arteries (ICAs) and the basilar artery (BA)
-- as root nodes, and designate the other ends as leaf nodes. The
resulting perivenous network graph consists of $\fixme{N}$ edges
connected at $\fixme{N}$ inner nodes and with $\fixme{N}$ end nodes,
and the associated domain is denoted by $\Lambda_v$. \fixme{Finally, we label
the outer surface of the PVS associated with $\Lambda_a$ and
$\Lambda_v$, by $\Gamma_a$ and $\Gamma_v$ respectively.}
